Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-11-09 analysis/paper/paper.qmd
Dealt with Ben's comments

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and  indirect bias. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. We found that some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more dependent on raw materials or other factors. I'M A BIT CONFUSED ABOUT THIS LAST SENTENCE, SEEMS LIKE IT SHOULD REFER TO SOCIAL CONTEXTS OF TRANSMISSION, BUT IT TALKS ABOUT RAW MATERIALS? CAN YOU LINK THE TWO?

@@

# Modeling the social context of the appearance of stemmed points in the Late Palaeolithic of Korea {.unnumbered}

@@

We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition . Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to facilitate observations of change over time.

@@

To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products. Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea, we selected roughly 25 as the threshold to decide between guided variation (\<25) and indirect bias (\>25). I THINK WE NEED TO EXPLAIN THIS DECISION OF THE THRESHOLD VALUE IN MORE DETAIL. I EXPECT REVIEWERS WILL HAVE QUESTIONS ABOUT WHY 25 IS THE NUMBER YOU CHOSE.

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Since our sample size is `r total_number_of_sp` artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results. [@kelley2007sample; @albatineh2014confidence]. I THINK THIS 'standard CV formula' IS A BAD CHOICE AND CONFUSING TO REVIEWERS. THIS IS BECAUSE IN SOME OF OUR PLOTS BELOW THERE ARE AS FEW AS 2 ARTEFACTS, THAT IS A VERY SMALL SAMPLE SIZE. GIVEN WHAT YOU HAVE WRITTEN ABOVE, WE SHOULD USE THE 'CORRECTED CV' BECAUSE MANY OF OUR COMPARISONS ARE WITH N\<25, FOR EXAMPLE, MOST SITES HAVE \<25 SP, SO WE *DO* HAVE SMALL SAMPLES AND WE NEED TO DEAL WITH IT USING THE BEST METHODS AVAILABLE.

@@

## Distribution of attributes

@fig-plot-cv-all-attributes shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points. This result implies a similar thickness of a wooden shaft that is attached to the stem.

```{r}
#| label: fig-plot-attributes

# Box plot for each attribute

# mutate data to make plot with ggplot
data_box <- gather(DF)

# make box plot for each attributes
plot_raw_measurements_for_attributes <- 
ggplot(data_box,
       aes(key, value)) +
  geom_boxplot() +
  geom_beeswarm(alpha = 0.3,
                size = 0.8,
                cex = 0.7) +
  ylab("Length (cm)") +
  xlab("") +
  theme_bw()
```

@@

@@

We used CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes (@fig-plot-cv-all-attributes panel A), the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques] and the relatively lower quality of lithic raw materials in Korea. If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.

The CV values for all stemmed points are mostly over 25, except for maximum length which is 23.8. PCA analysis shows positive relationships between attributes. We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time). After excluding two stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower. In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25. We observe the different degree of contribution by each attribute to the shape of stemmed points. For example, tang and stem related attributes have higher CV values compared to the body related attributes. We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.

@@

Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation. All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points. We assume that the guided variation played a more important role at the beginning of the transition based on high CV values of Phase 1. We observe a slight decrease in CV values over time. No clear regional patterns are confirmed. CV values could be vary depending on raw materials. For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals. Overall, we conclude that combination of the two transmission biases well explains the technological transition during the Late Paleolithic. We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission. We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks. In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.

>>>>

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and  indirect bias. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more associated to guided variation. Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group and then modified the shape according to their specific circumstances.

@@

# Modeling the social context of the appearance of stemmed points in the Late Paleolithic of Korea {.unnumbered}

@@

We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to facilitate observations of change over time.

@@

To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products. Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea and the median of the CV range (30.5), we selected roughly 25 as the threshold to decide between guided variation and indirect bias. We premised that CV values that are lower or similar to 25 represent guided variation while CV values over 25 reflect indirect bias.

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, we used the corrected CV formula since we computed CV per assemblages to examine the temporal and regional pattern. We also computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results. [@kelley2007sample; @albatineh2014confidence].

@@

@@

@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points. This result implies a similar thickness of a wooden shaft that is attached to the stem.

@@

We used the corrected CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes panel B). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

@@

To investigate the social contexts of technological transitions represented by the emergence of stemmed points during the Korean Late Paleolithic, we applied concepts of cultural transmission. We asked three questions to examine the dominant mode of cultural transmission and the transmission process over time: what is the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Do the modes of cultural transmission vary over time? and Do the modes of cultural transmission vary over space? We focused on two transmission biases: guided variation (socially isolated groups made stemmed points through trial and error) and indirect bias (socially connected groups whose knowledge of stemmed points derived from copying others). We set roughly 25 as our threshold of the CV value to decide between these two types of transmission biases, based on prior work on Great Basin Projectiles [@eerkens2001techniques]. If the CV value is lower than 25, which indicates indirect bias while the value is 25 or higher means guided variation.

The CV values for all stemmed points are mostly over 25, except for maximum length. PCA analysis shows positive relationships between attributes (@fig-plot-cv-all-attributes panel B). We then examined the CV values for chronological phases in the Korean Late Paleolithic period (@fig-plot-cv-over-time). After excluding two stemmed points for the Phase 1 to compute CV due to the small sample size issue, the overall CV values become lower. In addition to the maximum length, body length, mid width, and tang width for the Phase 2 are under 25. We observe the different degree of contribution by each attribute to the shape of stemmed points. For example, tang and stem related attributes have higher CV values compared to the body related attributes. We expect that people chose to emulate the body related attributes and had more freedom to produce the tang and stem related attributes.

@@

Our results show that some attributes such as body length, maximum length, mid width, and tang width were transmitted through indirect bias and the other attributes were developed through guided variation. All attributes are positively related and those low variation attributes are more closely associated and highly contribute to the overall shape of the stemmed points. We assume that the guided variation played a more important role at the beginning of the transition based on high CV values of Phase 1. We observe a slight decrease in CV values over time. No clear regional patterns are confirmed. CV values could be depending on raw materials. For example, coarse materials like quartzite have the highest CV values with the widest range of confidence intervals. Overall, we conclude that combination of the two transmission biases well explains the technological transition during the Late Paleolithic. We anticipate that our theoretical approach is applicable to other studies about technological transition and cultural transmission. We used landmarks to earn the dimensional attributes, which has potential for future research with different dimensions by calculating the distance between other sets of the existing landmarks. In our future research, we will focus on understanding actual connection between the assemblages and examining the social networks for sharing technology between the groups such as trades or physical migration to different locations.
